Icicle: write once, run once

### 2016-05-15 2016fhpc/section/03-Core.tex
some example query plans, some preliminary benchmark info

Plan  := query  x : T
       before { x : T = Exp... }
       folds  { x : T = Exp then Exp... }
       after  { x : T = Exp... }
       return { x : T = x... }

@@

Horizontal fusion for imperative loops is relatively easily, when the loops have the same number of iterations.
The problem with imperative loops occurs when trying to remove duplicate computations.
Firstly, the `definition' of a computation is split across multiple places: because the initial value of an accumulator and its update expression are in separate parts of the program, some analysis must be done to recover these.
For example, in the following loop, @a@ and @b@ have equivalent `update' expressions, but could denote very different computations depending on the implementation of @function@.
\begin{code}
a = 0;
b = 1;
for (...) {
  a = function(a);
  b = function(b);
}
\end{code}

Secondly, the order of statements in a loop affects the meaning: when accumulators are mutually dependent on each other, one can reorder the statements to produce a different, but still valid program.
These two loops are not equivalent:
\begin{code}
a = 0;
b = 1;
for (...) {
  a = a + b;
  b = a - b;
}
for (...) {
  b = a - b;
  a = a + b;
}
\end{code}

Neither of these problems are insurmountable, but the analysis required certainly complicates our goal of removing duplicate computation after fusion.

With this in mind, we will introduce our intermediate language based on what we call ``fold nests'', as opposed to loop nests.
Firstly, each accumulator should be defined in one place: this means that if two accumulators are alpha equivalent, they are equivalent.
Secondly, reordering statements (or accumulators) should not affect the meaning: two programs are equivalent if they contain equivalent sets of bindings, regardless of ordering.

We now convert the first example to folds:
\begin{code}
a = 0;
b = 1;
for (...) {
  a = function(a);
  b = function(b);
}

==>
loop ... {
  stage {
    fold a = 0
        then function(a);
  }
  stage {
    fold b = 1
        then function(b);
  }
}
\end{code}

Each fold binding must be grouped inside a @stage@ block, which allows mutual recursion across folds.
In this case, there is no mutual recursion, so the stages only contain one binding each.
The first fold is given the name @a@ with @fold a@.
Its initial value comes straight after the equals sign, so @a@ starts with a value of @0@.
The update or `kons' expression comes after @then@, and in the update expression any reference to @a@ refers to the current value of the fold.

It now becomes easier to see that the two folds are not equivalent, as the entire definitions are in one place.

The next example requires mutual recursion.
\begin{code}
a = 0;
b = 1;
for (...) {
  a = a + b;
  b = a - b;
}

==>
loop ... {
  stage {
    fold a = 0
        then a + b;
    fold b = 1
        then new a - b;
  }
}
\end{code}

Here, the update expression for @a@ refers to a later fold, @b@.
Update expressions can refer to later fold bindings in the same stage, as well as all earlier fold bindings.
The value referred to is the value of the binding at the start of the iteration, before any updates have occurred.
The new value of earlier folds, after being updated, can be referred to by suffixing a prime to the name: in the binding for @b@, the updated @a@ is referred to by @new a@.
This restriction of only accessing the new value of earlier folds is akin to the causality restriction in dataflow languages\cite{mandel2010lucy}.

Another way to think of the new value restriction above is disallowing cycles in the references to new values:
when @b@ references @new a@, the new value of @a@, it means that @new a@ must be computed before @new b@ can be.
If there were a cycle where @new a@ also depended upon @new b@, there would be no order to compute them in.
Disallowing cycles ensures an execution order.

By making the distinction between `old value' and `new value' explicit, the ordering in the program no longer has any effect on the semantics.
After converting the reordered loop with different semantics, the two programs are no longer alpha-equivalent:
\begin{code}
a = 0;
b = 1;
for (...) {
  b = a - b;
  a = a + b;
}

==>
loop ... {
  stage {
    fold b = 1
        then a - b;
    fold a = 0
        then a + new b;
  }
}
\end{code}

>>>>

Plan  := query  x with (x : T)...
       before { x : T = Exp... }
       folds  { x : T = Exp then Exp... }
       after  { x : T = Exp... }
       return { x : T = x... }

@@

One of the simplest queries to convert to a query plan is a count.
The query plan header specifies the table and the type of each column in the table.
This defines a single fold called @count@ which starts at zero, and is incremented for each element.
At the end of the query, the final count is assigned to the output @count_query@, which is the query's name.
\begin{code}
count_query =
query stocks
in    count

==>

query stocks with (open : Int) (close : Int)
before { }
folds  { count : Int = 0 then count + 1 }
after  { }
return { count_query : Int = count }
\end{code}

Computing the sum is only slightly more complicated.
Here there is again a single fold, @sum@, starting at zero and being incremented by the current column's @open@.
\begin{code}
sum_open =
query stocks
in    sum open

==>

query stocks with (open : Int) (close : Int)
before { }
folds  { sum : Int = 0 then sum + open }
after  { }
return { sum_open : Int = sum }
\end{code}

In order to compute the mean, we compute the sum and the count, and after all computation has finished, divide the two.
Here we use two folds, one for the sum, and one for the count.
While we could have used a single fold containing a pair of the two, keeping the folds separate exposes more opportunities for common subexpression elimination.
\begin{code}
mean_open =
query stocks
in    sum open / count

==>

query stocks with (open : Int) (close : Int)
before { }
folds  { sum       : Int = 0 then sum + open 
         count     : Int = 0 then count + 1 }
after  { mean      : Int = sum / count }
return { mean_open : Int = mean }
\end{code}

If we now wish to fuse these query plans together, it is as simple as giving the bindings unique names, and appending each part of the plan together.
The names of the returns and the column names stay the same.
Query plans can only be fused if they act on the same table.
\begin{code}
query stocks with (open : Int) (close : Int)
before { }
folds  { c1 : Int = 0 then c1 + 1
         c2 : Int = 0 then c2 + 1
         s1 : Int = 0 then s1 + open 
         s2 : Int = 0 then s2 + open }
after  { m1 : Int = s2 / c2 }
return { count_query : Int = c1
         sum_open    : Int = s1
         mean_open   : Int = m1 }
\end{code}

The above query plan computes all three queries at once, but has some duplicate work: the counts and the sums are performed twice.
We can perform fairly standard common subexpression elimination to remove these.
\begin{code}
query stocks with (open : Int) (close : Int)
before { }
folds  { c1 : Int = 0 then c1 + 1
         s1 : Int = 0 then s1 + open }
after  { m1 : Int = s1 / c1 }
return { count_query : Int = c1
         sum_open    : Int = s1
         mean_open   : Int = m1 }
\end{code}

The interesting thing here is just how simple fusion is.
Thanks to the single-pass restriction, we know there are no dependencies between any two query plans: if there were, the source query would have required multiple passes.
The single-pass restriction means that any source query can be expressed as a single fold over the data.
By keeping the folds separate we have exposed more opportunities for common subexpression elimination.

An efficient imperative loop can be generated from this query plan rather easily, in a similar way as previous work on flow fusion\cite{lippmeier2013data}.

### 2016-05-15 2016fhpc/section/04-Benchmarks.tex
some example query plans, some preliminary benchmark info

@@

>>>>

We are using Icicle in production, over medium-sized datasets that fit on a single disk.

One evaluation we performed was replacing a hand-written R script which had been running in production.
This R script works over around three hundred gigabytes of PSV data.
It computes twelve queries over each of the thirty-one input tables, computing 372 queries in total.

The R script for this takes around fifteen hours to run and is 3,566 lines, 2,311 of which are code.
In contrast, our Icicle queries take eleven minutes to run, and the dictionary describing the queries is 191 lines of code.

\begin{code}
          R       Icicle
Minutes   900         11
LOC       2,311      191
MB/s      6.6        543
\end{code}

@@

\begin{code}
        grep "^$"   wc    wc -l   Icicle
MB/s      74        251   775       1,060
\end{code}

### 2016-05-15 2016fhpc/section/XX-Elimination.tex
some example query plans, some preliminary benchmark info

>>>>

Horizontal fusion for imperative loops is relatively easily, when the loops have the same number of iterations.
The problem with imperative loops occurs when trying to remove duplicate computations.
Firstly, the `definition' of a computation is split across multiple places: because the initial value of an accumulator and its update expression are in separate parts of the program, some analysis must be done to recover these.
For example, in the following loop, @a@ and @b@ have equivalent `update' expressions, but could denote very different computations depending on the implementation of @function@.
\begin{code}
a = 0;
b = 1;
for (...) {
  a = function(a);
  b = function(b);
}
\end{code}

Secondly, the order of statements in a loop affects the meaning: when accumulators are mutually dependent on each other, one can reorder the statements to produce a different, but still valid program.
These two loops are not equivalent:
\begin{code}
a = 0;
b = 1;
for (...) {
  a = a + b;
  b = a - b;
}
for (...) {
  b = a - b;
  a = a + b;
}
\end{code}

Neither of these problems are insurmountable, but the analysis required certainly complicates our goal of removing duplicate computation after fusion.

With this in mind, we will introduce our intermediate language based on what we call ``fold nests'', as opposed to loop nests.
Firstly, each accumulator should be defined in one place: this means that if two accumulators are alpha equivalent, they are equivalent.
Secondly, reordering statements (or accumulators) should not affect the meaning: two programs are equivalent if they contain equivalent sets of bindings, regardless of ordering.

We now convert the first example to folds:
\begin{code}
a = 0;
b = 1;
for (...) {
  a = function(a);
  b = function(b);
}

==>
loop ... {
  stage {
    fold a = 0
        then function(a);
  }
  stage {
    fold b = 1
        then function(b);
  }
}
\end{code}

Each fold binding must be grouped inside a @stage@ block, which allows mutual recursion across folds.
In this case, there is no mutual recursion, so the stages only contain one binding each.
The first fold is given the name @a@ with @fold a@.
Its initial value comes straight after the equals sign, so @a@ starts with a value of @0@.
The update or `kons' expression comes after @then@, and in the update expression any reference to @a@ refers to the current value of the fold.

It now becomes easier to see that the two folds are not equivalent, as the entire definitions are in one place.

The next example requires mutual recursion.
\begin{code}
a = 0;
b = 1;
for (...) {
  a = a + b;
  b = a - b;
}

==>
loop ... {
  stage {
    fold a = 0
        then a + b;
    fold b = 1
        then new a - b;
  }
}
\end{code}

Here, the update expression for @a@ refers to a later fold, @b@.
Update expressions can refer to later fold bindings in the same stage, as well as all earlier fold bindings.
The value referred to is the value of the binding at the start of the iteration, before any updates have occurred.
The new value of earlier folds, after being updated, can be referred to by suffixing a prime to the name: in the binding for @b@, the updated @a@ is referred to by @new a@.
This restriction of only accessing the new value of earlier folds is akin to the causality restriction in dataflow languages\cite{mandel2010lucy}.

Another way to think of the new value restriction above is disallowing cycles in the references to new values:
when @b@ references @new a@, the new value of @a@, it means that @new a@ must be computed before @new b@ can be.
If there were a cycle where @new a@ also depended upon @new b@, there would be no order to compute them in.
Disallowing cycles ensures an execution order.

By making the distinction between `old value' and `new value' explicit, the ordering in the program no longer has any effect on the semantics.
After converting the reordered loop with different semantics, the two programs are no longer alpha-equivalent:
\begin{code}
a = 0;
b = 1;
for (...) {
  b = a - b;
  a = a + b;
}

==>
loop ... {
  stage {
    fold b = 1
        then a - b;
    fold a = 0
        then a + new b;
  }
}
\end{code}
